Evolution of pre-copulatory and post-copulatory strategies of inbreeding avoidance

### 2017-07-28 ms_pre_post.Rmd
Substantial changes to end of Discussion

In one first step, @Duthie used a genetically-explicit individual-based model to examine conditions under which polyandry is predicted to evolve due to selection stemming from pre-copulatory inbreeding avoidance in the absence of post-copulatory inbreeding avoidance. Simulations showed that even when selection for pre-copulatory inbreeding avoidance was strong and females consequently preferred unrelated mates, selection for polyandry specifically to facilitate this inbreeding avoidance occurred only under highly restricted conditions. Key requirements were that direct negative selection (i.e., 'costs') on polyandry was weak, that very few males were available for a female's initial mate choice but many were available for additional mate choice(s), or that polyandry was conditionally expressed when a focal female was related to her initial mate [@Duthie]. Without these conditions, polyandrous females tended to increase rather than decrease their overall degree of inbreeding, ultimately reducing offspring fitness. This increase occurred because, once pre-copulatory inbreeding avoidance evolved, polyandrous females had already chosen available unrelated males as their initial mates. Their additional mates, chosen from the remaining male population, were therefore increasingly likely to include relatives. Evolution of polyandry purely to facilitate pre-copulatory inbreeding avoidance, and hence resulting inbreeding avoidance, was consequently restricted [@Duthie].

@@

Offsprings' alleles mutate with independent probabilities $\mu=0.001$. When a mutation occurs, a mutation effect is sampled from a normal distribution with a mean of zero and a standard deviation of $\mu_{SD}$ and added to the original allele value [@Kimura1965; @Lande1976; @Bocedi2014; @Duthie]. The value of $\mu_{SD}$ is set to $\sqrt{1/20}$.

@@

The viability of a female $i$'s offspring ($\Psi_{\textrm{off}}$) decreases as a log-linear function of her kinship with the sire $j$ of the offspring ($k_{ij}$) and inbreeding depression slope $\beta$,
\begin{equation}
\Psi_{\textrm{off}} = e^{-\beta k_{ij}}
\end{equation}
Here, $\beta$ models the number of haploid lethal equivalents that exist as deleterious recessive alleles in the gametes of $i$ and $j$, and which might be homozygous in offspring and reduce viability. Equation 1 assumes independent allelic effects, generating multiplicative effects on offspring viability [@Morton1956; @Mills1994]. It also assumes that inbreeding does not covary with inbreeding load (i.e., no purging). This formulation ensures that the relationship between $k_{ij}$ and the magnitude of inbreeding depression in offspring is consistent across replicate simulations and is justified because previous genetically-explicit modelling [@Duthie2016a] showed that inbreeding avoidance in biparental populations has a negligible effect on load given small-effect deleterious mutations [see also @Wang1999; @Guillaume2006]. We model inbreeding depression as having an absolute rather than relative effect on offspring viability (i.e., hard rather than soft selection), so that the effect of $\beta$ is consistent across generations and different parameter combinations. We assume that inbreeding always decreases offspring viability (i.e., $\beta > 0$, giving inbreeding depression but no outbreeding depression). Therefore, because $0 \leq k_{ij} \leq 1$, $-\beta \times k_{ij} \leq 0$. Values of $\Psi_{\textrm{off}}$ must therefore be between zero (if $-\beta \times k_{ij}$ is very negative) and one (if $-\beta \times k_{ij} = 0$). We therefore define $\Psi_{\textrm{off}}$ as the probability that an offspring is viable, and sample its realised viability (versus mortality) using a Bernoulli trial. Offspring that are viable after the Bernoulli trial become adults. Given our current objectives, we simulate evolution under conditions where inbreeding avoidance is adaptive due to strong inbreeding depression, not where inbreeding preference is adaptive due to weak or zero inbreeding depression [@Parker1979; @Kokko2006; @Duthie2016a] or outbreeding depression [@Bateson1983; @Greeff2009]. However, as described above, positive $M_{p}$ and $F_{p}$ values resulting in inbreeding preference are not precluded from evolving, and could potentially arise due to mutation or drift.

@@

We used individual-based modelling to highlight fundamental but theoretically under-developed relationships between evolution of polyandry and pre-copulatory versus post-copulatory inbreeding strategy given (1) hard constraints on evolution of post-copulatory inbreeding strategy, (2) asymmetric costs of pre-copulatory and post-copulatory inbreeding strategy, and (3) evolution of one inbreeding strategy phenotype given pre-existence of the other. We present our model and simulation results as tools for thinking more clearly about the dynamics of simultaneously evolving reproductive strategies, not for making accurate quantitative predictions for any specific system. While the logic of our model can be usefully applied to construct hypotheses within and across empirical systems, accurate quantitative prediction would require additional empirical detail and data for model parameterisation.

General predictions concerning inbreeding avoidance and polyandry
-------------------------------------------------------------------

Post-copulatory inbreeding avoidance cannot be effectively realised if females are not polyandrous in any form, and is likely to be most effective for highly polyandrous females that can bias fertilisation amongst sperm contributed by multiple mates. In contrast, pre-copulatory inbreeding avoidance mechanisms are most critical for females that mate only once and therefore have no post-copulatory opportunity to avoiding inbreeding. This theory is borne out in our simulation results, as selection for, and consequent evolution of, post-copulatory inbreeding avoidance was negligible in populations where polyandry did not evolve, resulting in high drift of allele values over generations due to the inability of females to express post-copulatory choice (e.g., Figure 4C). Evolution of pre-copulatory inbreeding avoidance was also typically slightly stronger when polyandry did not evolve (e.g., Figure 3A vs 3B; see also Supporting Information Figure S11). In addition to initial polyandry causing evolution of post-copulatory inbreeding avoidance, polyandry might also covary positively with post-copulatory inbreeding avoidance due to the feed back effect that post-copulatory inbreeding avoidance has on facilitating evolution of polyandry itself, as observed in our model (Figure 1). It would therefore be interesting to test the hypothesis that, across taxa, the occurrence of post-copulatory inbreeding avoidance covaries positively, and the occurrence of pre-copulatory inbreeding avoidance covaries negatively, with the degree of polyandry. To test this hypothesis, more work is needed to quantify the degree to which females of different species engage in polyandry and the degree to which females express both pre-copulatory and post-copulatory inbreeding avoidance.

@@

Further, pre-existence of fixed adaptive pre-copulatory inbreeding avoidance precluded evolution of polyandry and, in turn, precluded evolution of post-copulatory inbreeding avoidance (Figure 3C,D). However, existence of fixed adaptive post-copulatory inbreeding avoidance did not preclude evolution of pre-copulatory inbreeding avoidance (Figure 3A,B). In most populations, it is unlikely that pre-copulatory and post-copulatory inbreeding avoidance will evolve simultaneously from an ancestral population in which females mate and assign paternity randomly. Rather, the timing of the invasion of adaptive inbreeding avoidance phenotypes will differ, so the initial evolution of one inbreeding strategy will likely occur in the absence of the other, or where selection for and subsequent evolution of the other strategy has already occurred. When framing hypotheses for existence of post-copulatory inbreeding avoidance and polyandry, it might therefore be necessary to consider whether or not inbreeding avoidance already occurs through pre-copulatory mate choice. Additionally, the opportunity for post-copulatory inbreeding avoidance will also depend on the degree to which females are polyandrous. For species in which pre-copulatory inbreeding avoidance occurs and polyandry is uncommon [@Lihoreau2007; @Metzger2010; @Metzger2010a], evolution of post-copulatory inbreeding avoidance is unlikely even if such a strategy incurs little direct cost.

@@

To acheive our current objective of providing general conceptual comparison of the effects of direct costs across phenotypes, we modelled all costs as analogous increased probabilities of total reproductive failure, thereby facilitating direct comparison across traits. This cost formulation reflects empirical observations in some populations [see Model; e.g., @Rowe1994; @Koga1998; @Gasparini2011], and is therefore a biologically realistic means of standardising costs across traits. However, different forms of costs could be incorporated into future models designed to predict evolutionary dynamics that might occur in specific systems. 
@Pomiankowski1987 identified four costs categories relevant to mating frequency and mate choice, including elevated risks of predation or disease transmission, and time or energey expenditure. For example, polyandrous females might experience increased risk of disease transmission [@Roberts2015]. This cost would more realistically apply to a female's realised number of mates rather than her tendency for polyandry. Polyandrous females might also risk harm caused by sexual conflict over multiple mating [e.g., @Arnqvist2005; @Parker2006]. Inbreeding theory predicts that males should be more tolerant of inbreeding females, leading to sexual conflict over inbreeding [@Parker1979; @Parker2006; @Kokko2006; @Duthie2015a]. Future models could therefore explicitly consider sexual conflict over both polyandry and inbreeding, and hence capture internally-consistent mechanistic costs.

<!--- TODO: Edit first sentence --->
Models could then be further developed such that costs arise from explicit reproductive mechanisms, requiring further biological detail. Recent studies have elucidated key mechanisms of post-copulatory inbreeding avoidance. For example, when female guppies (*Poecilia reticulata*) were artificially inseminated with equal quantitites of sperm from full-siblings and unrelated males, more eggs were fertilised by unrelated males because the velocities of full sibling's sperm were reduced by females' ovarian fluids [@Gasparini2011]. In black field crickets (*Teleogryllus commodus*), females attempt to remove the spermatophores of unwanted males after copulation, and are capable of controlling sperm transfer to spermatheca after copulation occurs [@Bussiere2006; @Tuni2013].

<!--- TODO: Edit first sentence ---> 
We restricted our model to simulate how evolution of traits were affected by their direct costs and indirect benefits due to inbreeding avoidance. Although empirical populations will include additional beneficial traits that might affect evolution of mating strategies, targeting our focus on traits affecting inbreeding and polyandry allows us to better isolate their effects and therefore better develop hypotheses for natural systems.

>>>>

In one first step, @Duthie used a genetically-explicit individual-based model to examine conditions under which polyandry is predicted to evolve due to selection stemming from pre-copulatory inbreeding avoidance in the absence of post-copulatory inbreeding avoidance. Simulations showed that even when selection for pre-copulatory inbreeding avoidance was strong and females consequently preferred unrelated mates, selection for polyandry specifically to facilitate this inbreeding avoidance occurred only under highly restricted conditions. Key requirements were that direct negative selection (i.e., 'costs') on polyandry was weak, that very few males were available for a female's initial mate choice but many were available for additional mate choice(s), or that polyandry was conditionally expressed when a focal female was related to her initial mate [@Duthie]. Without these conditions, polyandrous females tended to increase rather than decrease their overall degree of inbreeding, ultimately reducing offspring fitness. This increase occurred because, once pre-copulatory inbreeding avoidance evolved, polyandrous females had already chosen available unrelated males as their initial mates. Their additional mates, chosen from the remaining male population, were therefore increasingly likely to include relatives. Evolution of polyandry purely to facilitate pre-copulatory inbreeding avoidance was consequently restricted [@Duthie].

@@

Offsprings' alleles mutate with independent probabilities $\mu=0.001$. When a mutation occurs, a mutation effect size is sampled from a normal distribution with a mean of zero and a standard deviation of $\mu_{SD}$ and added to the original allele value [@Kimura1965; @Lande1976; @Bocedi2014; @Duthie]. The value of $\mu_{SD}$ is set to $\sqrt{1/20}$.

@@

The viability of a female $i$'s offspring ($\Psi_{\textrm{off}}$) decreases as a log-linear function of her kinship with the sire $j$ of the offspring ($k_{ij}$) and inbreeding depression slope $\beta$,
\begin{equation}
\Psi_{\textrm{off}} = e^{-\beta k_{ij}}
\end{equation}
Here, $\beta$ models the number of haploid lethal equivalents that exist as deleterious recessive alleles in the gametes of $i$ and $j$, and which might be homozygous in offspring and reduce viability. Equation 1 assumes independent allelic effects, generating multiplicative effects on offspring viability [@Morton1956; @Mills1994]. It also assumes that inbreeding does not covary with inbreeding load (i.e., no purging). This formulation ensures that the relationship between $k_{ij}$ and the magnitude of inbreeding depression in offspring is consistent across replicate simulations and is further justified because previous genetically-explicit modelling [@Duthie2016a] showed that inbreeding avoidance in biparental populations has a negligible effect on load given small-effect deleterious mutations [see also @Wang1999; @Guillaume2006]. We model inbreeding depression as having an absolute rather than relative effect on offspring viability (i.e., hard rather than soft selection) so that the effect of $\beta$ is consistent across generations and different parameter combinations. We assume that inbreeding always decreases offspring viability (i.e., $\beta > 0$, giving inbreeding depression but no outbreeding depression). Therefore, because $0 \leq k_{ij} \leq 1$, $-\beta \times k_{ij} \leq 0$. Values of $\Psi_{\textrm{off}}$ must therefore be between zero (if $-\beta \times k_{ij}$ is very negative) and one (if $-\beta \times k_{ij} = 0$). We therefore define $\Psi_{\textrm{off}}$ as the probability that an offspring is viable, and sample its realised viability (versus mortality) using a Bernoulli trial. Offspring that are viable after the Bernoulli trial become adults. Given our current objectives, we simulate evolution under conditions where inbreeding avoidance is adaptive due to strong inbreeding depression, not where inbreeding preference is adaptive due to weak or zero inbreeding depression [@Parker1979; @Kokko2006; @Duthie2016a] or outbreeding depression [@Bateson1983; @Greeff2009]. However, as described above, positive $M_{p}$ and $F_{p}$ values resulting in inbreeding preference are not precluded from evolving, and could potentially arise due to mutation or drift.

@@

We used individual-based modelling to highlight fundamental but theoretically under-developed relationships between evolution of polyandry and pre-copulatory versus post-copulatory inbreeding strategy given (1) hard constraints on evolution of post-copulatory inbreeding strategy, (2) asymmetric costs of pre-copulatory and post-copulatory inbreeding strategy, and (3) evolution of one inbreeding strategy phenotype given pre-existence of the other. We present our model and simulation results as tools for thinking more clearly about the dynamics of simultaneously evolving reproductive strategies, not for making accurate quantitative predictions for any specific system. While the logic of our model can be usefully applied to construct general hypotheses within and across empirical systems, accurate quantitative prediction would require additional empirical detail and data for model parameterisation.

General hypotheses concerning inbreeding avoidance and polyandry
-------------------------------------------------------------------

Post-copulatory inbreeding avoidance cannot be effectively realised if females are not polyandrous in any form, and is likely to be most effective for highly polyandrous females that can bias fertilisation amongst sperm contributed by multiple mates. In contrast, pre-copulatory inbreeding avoidance mechanisms are most critical for females that mate only once and therefore have no post-copulatory opportunity to avoid inbreeding. This theory is borne out in our simulation results, as selection for, and consequent evolution of, post-copulatory inbreeding avoidance was negligible in populations where polyandry did not evolve, resulting in high drift of allele values over generations due to the inability of females to express post-copulatory choice (e.g., Figure 4C). Evolution of pre-copulatory inbreeding avoidance was also typically slightly stronger when polyandry did not evolve (e.g., Figure 3A vs 3B; see also Supporting Information Figure S11). In addition to initial polyandry causing evolution of post-copulatory inbreeding avoidance, polyandry might also covary positively with post-copulatory inbreeding avoidance due to the feed back effect that post-copulatory inbreeding avoidance has on facilitating evolution of polyandry itself, as observed in our model (Figure 1). It would therefore be interesting to test the hypothesis that, across taxa, the occurrence of post-copulatory inbreeding avoidance covaries positively, and the occurrence of pre-copulatory inbreeding avoidance covaries negatively, with the degree of polyandry. To test this hypothesis, more work is needed to quantify the degree to which females of different species engage in polyandry and the degree to which females express both pre-copulatory and post-copulatory inbreeding avoidance.

@@

Further, pre-existence of fixed adaptive pre-copulatory inbreeding avoidance precluded evolution of polyandry and, in turn, precluded evolution of post-copulatory inbreeding avoidance (Figure 3C,D). However, pre-existence of fixed adaptive post-copulatory inbreeding avoidance did not preclude evolution of pre-copulatory inbreeding avoidance (Figure 3A,B). In natural populations, it is unlikely that pre-copulatory and post-copulatory inbreeding avoidance will evolve simultaneously from an ancestral population in which females mate and assign paternity randomly. Rather, the timing of the invasion of adaptive inbreeding avoidance phenotypes will differ, so the initial evolution of one inbreeding strategy will likely occur in the absence of the other, or where selection for and subsequent evolution of the other strategy has already occurred. When framing hypotheses for existence of post-copulatory inbreeding avoidance and polyandry, it might therefore be necessary to consider whether or not inbreeding avoidance already occurs through pre-copulatory mate choice. Additionally, the opportunity for post-copulatory inbreeding avoidance will also depend on the degree to which females are polyandrous. For species in which pre-copulatory inbreeding avoidance occurs and polyandry is uncommon [@Lihoreau2007; @Metzger2010; @Metzger2010a], evolution of post-copulatory inbreeding avoidance is unlikely even if such a strategy incurs little direct cost.

@@

To achieve our current objective of providing general conceptual comparison of the effects of direct costs across phenotypes, we modelled all costs as analogous increased probabilities of total reproductive failure. This cost formulation reflects empirical observations in some populations [see Model; e.g., @Rowe1994; @Koga1998; @Gasparini2011], and is therefore a biologically realistic means of standardising costs across traits. However, different forms of costs could be incorporated into future models designed to predict evolutionary dynamics that might occur in specific systems. Models could then be further developed such that costs arise from explicit reproductive mechanisms, requiring further biological detail.

For example, @Pomiankowski1987 identified four costs categories relevant to mating frequency and mate choice, including elevated risks of predation or disease transmission, and time or energey expenditure. For example, polyandrous females might experience increased risk of disease transmission [@Roberts2015]. This cost would more realistically apply to a female's realised number of mates rather than her tendency for polyandry. Polyandrous females might also risk harm caused by sexual conflict over multiple mating [e.g., @Arnqvist2005; @Parker2006]. Inbreeding theory predicts that males should be more tolerant of inbreeding than females, leading to sexual conflict over inbreeding in mating encounters [@Parker1979; @Parker2006; @Kokko2006; @Duthie2015a]. Future models could therefore explicitly consider sexual conflict over both polyandry and pre-copulatory inbreeding, and hence capture internally-consistent mechanistic costs. Further, sexual conflict might also affect post-copulatory inbreeding. For example, when female guppies (*Poecilia reticulata*) were artificially inseminated with equal quantitites of sperm from full-siblings and unrelated males, more eggs were fertilised by unrelated males because the velocities of full sibling's sperm were reduced by females' ovarian fluids [@Gasparini2011]. In black field crickets (*Teleogryllus commodus*), females attempt to remove the spermatophores of unwanted males after copulation, and are capable of controlling sperm transfer to spermatheca after copulation occurs [@Bussiere2006; @Tuni2013].

We restricted our model to simulate how evolution of traits was affected by their direct costs and indirect benefits due to inbreeding avoidance. Although natural populations will include additional beneficial traits that might affect evolution of mating strategies, targeting our focus on traits affecting inbreeding and polyandry allows us to better isolate their effects and therefore better develop general hypotheses. Future models could be further developed to include additional benefits and costs of reproductive strategies, and benefit and cost values could be parameterised using data from empirical populations, hence facilitating more quantitative predictions for specific systems. The addition of such mechanistic detail, and the inclusion of data from populations of interest to parameterise model extensions, could ultimately improve understanding and prediction of reproductive strategy evolution and thereby link theory, modelling, and empirical hypothesis testing.
